---
title: "BumbleboxExampleAnalysis"
output: pdf_document
date: "2023-11-07"
---

Written by James Roberts Crall (james.crall@wisc.edu) for the BOMBUSS 3.0 tracking workshop

November 13, 2023

Code available at: www.github.com/Crall-Lab/BumbleBox_BOMBUSS

# Set up
```{r setup}
#Load packages
library(ggplot2)
library(png)
library(lubridate)
library(stringr)
library(grid)
library(lme4)
library(lmerTest)
library(gridExtra)
library(igraph)

#Define parent directory containing data you want to analyze
#This example uses data from a single Bombus impatiens microcolony (data will be provided to workshop participants)

pdir <- '/Users/jamescrall/Downloads/2023-10-28'
brood_metadata <- read.csv('/Users/jamescrall/Downloads/2023-10-28/bumblebox-01-10-28-nest_image.csv')

#Define image dimensions for plotting
xl <- 4056
yl <- 3040
```

# Load and plot data tracking data from single video
```{r load example data set for one video}

#Get list of unique trials (using recursive search in case there are subfolders)
trials <- list.files(pdir, pattern = '*averages.csv', full.names = TRUE, recursive = TRUE)

#Select a random trial - trial 10 in this list
i <- 10

#Load background photo
background_im <- readPNG(str_replace(trials[i], '_averages.csv', '.png'))

#Clip image excess
background_im <- background_im[1:3040,]

#Optional: plot image (this can be slow)
#image(background_im, col = gray.colors(33))


#load and plot tag tracking data
tracking_data <- read.csv(str_replace(trials[i], '_averages.csv', '_updated.csv'))
tracking_data$ID <- as.factor(tracking_data$ID)
ggplot(tracking_data, aes(x = centroidX, y = centroidY, colour = ID)) + geom_point()+xlim(c(0,xl)) + ylim(c(0,yl))

#average data
summary_data <- read.csv(trials[i])
head(summary_data)
```



# Explore data across all trials

### First loop and compile across individual tracking files to compile into a single data frame
```{r, compile all raw tracking data}
#Make sure you don't have a 'comb_data' object in memory
rm('tracking_data')

for(file in trials){
  
  data <- read.csv(str_replace(file, '_averages.csv', '_updated.csv'))
  ## Compile average data
  #Check to see if output data frame exists
  if(!exists('tracking_data')){
    tracking_data <- data #If not, create it from the first loop
  } else {
    tracking_data <- rbind(tracking_data, data) #Otherwise append new data to the end
  }
  
}

tracking_data$ID <- as.factor(tracking_data$ID)
ggplot(tracking_data, aes(x = centroidX, y = centroidY, colour = ID)) + geom_point(alpha = 0.3, cex = 0.4)+xlim(c(0,xl)) + ylim(c(0,yl))

```
### Next, compile data on within-trial average data (generated directly from BumbleBox scripts)
```{r combine data on average metrics within videos}

#Make sure you don't have a 'comb_data' object in memory
rm('comb_data')

for(file in trials){
  
  data <- read.csv(file)
  ## Compile average data
  
  #Check to see if output data frame exists
  if(!exists('comb_data')){
    comb_data <- data #If not, create it from the first loop
  } else {
    comb_data <- rbind(comb_data, data) #Otherwise append new data to the end
  }
  
}

#Convert ID to factor instead of numeric
comb_data$ID <- as.factor(comb_data$ID)
```

### Plot variation in speed and distance from center across individuals
```{r plot individual data}

#Clean out missing data
comb_data <- comb_data[complete.cases(comb_data),]

#Boxplot of speed vs. individual
ggplot(comb_data, aes(x = reorder(ID, average.speed, FUN = median), y = log10(average.speed)))+geom_boxplot()

#Check for evidence of individual variation
model <- lmer(log10(1+average.speed)~ID + (1|filename), data = comb_data)
model.1 <- lmer(log10(1+average.speed)~1 + (1|filename), data = comb_data)
anova(model, model.1)

#Show distance from center by individual
ggplot(comb_data, aes(x = reorder(ID, average.distance.from.center, FUN = median), y = average.distance.from.center))+geom_boxplot()

#Get individual averages
individual_averages <- aggregate(cbind(average.distance.from.center, average.speed, frames.tracked.in.video)~ID, data = comb_data, FUN = median)
colnames(individual_averages) <- c('ID', 'dist', 'speed', 'frames_tracked')

# Remove outlier with zero velocity
individual_averages <- subset(individual_averages, speed > 1)

```
### Determine activity (moving vs. not moving) by thresholding speed
```{r movement thresholding}
#Plot histogram of framewise speeds to determine threshold for 'real' movement
hist(log10(tracking_data$speed), breaks = 30)

#Looks like ~3.1 pixels/frame (log10(3.1)~= 0.5)
speed_thresh <- 3.1
tracking_data$moving[tracking_data$speed > speed_thresh] <- 1
tracking_data$moving[tracking_data$speed < speed_thresh] <- 0

```

### Visualize changes in individual behavior over time
```{r, visualize individual behavior over time}
#Visualize individual speed/distance over time

#Set up timestamps
#Reference start time
start.date <- parse_date_time('2023-10-28 00:00:00', "%Y-%m-%d %H:%M:%S")

#Create timestamp variable
comb_data$datetime <- parse_date_time(substr(comb_data$filename, 14, 32), "%Y-%m-%d_%H-%M-%S")

#Convert into 'days since the start time'
comb_data$time.num <- as.numeric(difftime(comb_data$datetime, start.date, units = 'days'))

#Reorder
comb_data <- comb_data[order(comb_data$ID, comb_data$time.num),]

#Plots over time and individual variation for speed
p1 <- ggplot(comb_data, aes(x = time.num, y = average.speed, colour = ID)) + geom_point(alpha = 0.3, cex = 0.4)+geom_line(alpha = 0.3)
p2 <- ggplot(comb_data, aes(x = reorder(ID, average.speed, FUN = median), y = average.speed, colour = ID))+geom_boxplot()
grid.arrange(p1, p2, ncol=2)

#Plots over time and individual variation for distance from center
p1 <- ggplot(comb_data, aes(x = time.num, y = average.distance.from.center, colour = ID)) + geom_point(alpha = 0.3, cex = 0.4)+geom_line(alpha = 0.3)
p2 <- ggplot(comb_data, aes(x = reorder(ID, average.distance.from.center, FUN = median), y = average.distance.from.center, colour = ID))+geom_boxplot()
grid.arrange(p1, p2, ncol=2)

#Get average time spent moving for each bee
aggregate(moving~ID, data = tracking_data, FUN = mean)
```

### Plot individual spatial occupancy patterns within the nest
```{r spatial variation of individuals}
#Plot for single representative individual, with brood locations overlaid
ggplot(subset(tracking_data, ID == 9), aes(x = centroidX, y = centroidY)) + geom_density_2d_filled()+ggtitle('worker 9')+geom_point(data=brood_metadata, aes(x = x, y = y, colour = label))

#Plot spatial distribution for select individuals (representing centrality range)
#NB: if you are working on your own data set, select the IDs here from your own taglist
worker_list <- c('1', '13', '15', '3')
p1 <- ggplot(subset(tracking_data, ID == worker_list[1]), aes(x = centroidX, y = centroidY)) + geom_density_2d_filled(show.legend = FALSE)+ggtitle(paste('worker', worker_list[1], sep = " "))+geom_point(data=brood_metadata, aes(x = x, y = y, colour = label, alpha = 0.7), show.legend = FALSE)
p2 <- ggplot(subset(tracking_data, ID == worker_list[2]), aes(x = centroidX, y = centroidY)) + geom_density_2d_filled(show.legend = FALSE)+ggtitle(paste('worker', worker_list[2], sep = " "))+geom_point(data=brood_metadata, aes(x = x, y = y, colour = label, alpha = 0.7), show.legend = FALSE)
p3 <- ggplot(subset(tracking_data, ID == worker_list[3]), aes(x = centroidX, y = centroidY)) + geom_density_2d_filled(show.legend = FALSE)+ggtitle(paste('worker', worker_list[3], sep = " "))+geom_point(data=brood_metadata, aes(x = x, y = y, colour = label, alpha = 0.7), show.legend = FALSE)
p4 <- ggplot(subset(tracking_data, ID == worker_list[4]), aes(x = centroidX, y = centroidY)) + geom_density_2d_filled(show.legend = FALSE)+ggtitle(paste('worker', worker_list[4], sep = " "))+geom_point(data=brood_metadata, aes(x = x, y = y, colour = label, alpha = 0.7), show.legend = FALSE)

grid.arrange(p1,p2,p3,p4, ncol = 2)
```


### Analyze contact-based social networks. *NB: this code set up for a data set with sixteen workers, and would need to be updated for different colony sizes
```{r contact network analysis}

#Start by loading contact data by looping over source files

#Create empty output data structure
out_data <- as.data.frame(matrix(ncol = 17, nrow = 0))
colnames(out_data) <- as.character(c('ID', as.character(seq(0,15))))

#loop over trials
for(file in trials){
  
  #Get filename specifically for 'contacts.csv' data
  data <- read.csv(str_replace(file, '_averages.csv', '_contacts.csv'))
  
  #Get rid of 'X' in filenames
  names(data) <- gsub("X", "", names(data))
  
  #Remove columns that aren't in output data
  data <- data[,colnames(data) %in% colnames(out_data)]

  #Create and populate dummy intermediate data structure to align columns with output
  empty_data <- as.data.frame(matrix(ncol = 17, nrow = length(data[,1])))
  colnames(empty_data) <- colnames(out_data)
  empty_data[,match(colnames(data), colnames(out_data))] <- data
  
  #Aggregate data
  out_data <- rbind(out_data, empty_data)
  
}

#Now, get average interaction rates across all pairwise individuals
interactions <- aggregate(out_data[,2:17], by = list(out_data$ID), FUN = mean, na.rm = TRUE)

#Remove data from errant tags and restructure into adjacency matrix
interactions <- interactions[1:16,2:17]
interactions <- as.matrix(interactions)
interactions[is.na(interactions)] <- 0
interactions[interactions == 1] <- 0

#Make fructhermann-reingold graph using adjaceny matrix
net <- graph.adjacency(interactions, mode = 'undirected', weighted=TRUE, diag=FALSE)
plot.igraph(net,layout=layout.fruchterman.reingold, edge.width=E(net)$weight*50, vertex.size = 20)

```